---
title: "An introduction to `whotools`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An introduction to `whotools`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4,
  warning = FALSE
)
```

## Overview

`whotools` is intended as a lightweight package for facilitating analytics at
WHO. The following vignette will cover some example use cases. We will use the
example dataset `incidence`, containing daily reported cases and deaths for a
given country:

```{r setup}
## load packages
pacman::p_load(tidyverse, lubridate, glue, whotools, gt)

## load example data from package
incidence <- whotools:::incidence
```

## Visualisation

### Themes

`whotools` includes a theme that can be added to any ggplot2 object using the
`theme_who()` function. In the example below we are visualising the daily case
incidence curve with the WHO theme, also using the `scale_y_epicurve` function
to generate a y-axis appropriate for daily case data.

```{r, theme_who}
incidence %>%
  ggplot(aes(date, cases)) +
  geom_col(width = 1) +
  scale_y_epicurve("cases") +
  labs(x = "Date", y = "Daily number of cases") +
  theme_who()
```

### Palettes

The `who_palette` function provides a number of palettes that can be applied in
different scenarios and passed via the `scale_manual` option in ggplot2. In the
example below, we visualise the change in urban and non-urban populations in
Afghanistan over time using the `binary` palette.

```{r, who_palette_binary}
## load world bank population data from {tidyr} package
data("world_bank_pop", package = "tidyr")

## format data
world_bank_pop %>%
  filter(country == "AFG", indicator %in% c("SP.URB.TOTL", "SP.POP.TOTL")) %>%
  pivot_longer(
    -c(country, indicator),
    names_to = "year",
    values_to = "population"
  ) %>%
  pivot_wider(names_from = indicator, values_from = population) %>%
  transmute(country, year, urban = SP.URB.TOTL, non_urban = SP.POP.TOTL - SP.URB.TOTL) %>%
  pivot_longer(-c(country, year)) %>%
  ggplot(aes(year, value/1e6, fill = name)) +
  geom_col(position = "stack") +
  scale_fill_manual(
    ## use binary palette type
    values = who_palette(type = "binary2"),
    labels = c(urban = "Urban", non_urban = "Non-urban"),
    name = NULL
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0))) +
  guides(fill = guide_legend(reverse = TRUE)) +
  labs(
    x = "Year", y = "Population (millions)",
    title = "Urban and non-urban population growth in Afghanistan"
  ) +
  theme_who() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )  
```

### Region labels and palettes

The `relabel_who` function helps convert regional offices labels between their
abbreviated and full version (.e.g "EMRO" and "Regional Office for the Eastern
Mediterranean") and between office and region designations (e.g. EMRO and
EMR). Below we show average case incidence per region in a table, also using the
`theme_who_gt` function, which applies a WHO theme to table outputs from the
`gt` package.

```{r, region_labels}
cases_region <- tibble(
  office = c("AFRO", "EURO", "EMRO", "PAHO", "SEARO", "WPRO"),
  region = relabel_who(office, output = "region", length = "long"),
  cases = round(runif(length(office), 20, 200), 1.1)
)

cases_region %>%
  select(-office) %>%
  gt() %>%
  cols_label(cases = "Daily cases per 100,000 population", region = "Region") %>%
  tab_options(table.font.size = 12) %>%
  theme_who_gt()
```

We can also plot this in a bar chart and use the regional color palette from `who_palette`:

```{r, region_palette}
cases_region %>%
  mutate(office = fct_reorder(office, cases)) %>%
  ggplot(aes(office, cases, fill = office)) +
  geom_col() +
  scale_x_discrete(labels = ~ relabel_who(.x, output = "region", length = "short")) +
  scale_fill_manual(values = who_palette(type = "region"), guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "WHO Region", y = "Daily cases per 100,000 population") +
  theme_who()
```

## Methods

### Moving averages

The `moving_average` function calculates a moving average across a specified
date window. Below we add a line showing the average daily incidence over a
sliding 7-day window.

```{r, moving_average}
incidence %>%
  mutate(
    cases_weekly = moving_average(val = cases, date_val = date, before = 3, after = 3),
  ) %>%
  ggplot(aes(x = date)) +
  geom_col(aes(y = cases), width = 1) +
  geom_line(aes(y = cases_weekly), size = 2, color = "black") +
  scale_y_epicurve("incidence") +
  labs(x = "Date", y = "Daily number of cases") +
  theme_who()
```

### Percent changes

The `perc_change` function will calculate percent changes. We can use this to
visualise percent changes in weekly reported cases:

```{r, perc_change}
incidence %>%
  group_by(week = isoweek(date)) %>%
  filter(week < 30) %>%
  summarise(cases = sum(cases)) %>%
  mutate(change = 100*perc_change(cases, lag(cases))) %>%
  pivot_longer(-week) %>%
  ggplot(aes(week, value)) +
  geom_col() +
  facet_wrap(
    ~ name,
    ncol = 1,
    scales = "free_y",
    labeller = labeller(name = c(cases = "Weekly number\nof cases", change = "% change in\nweekly cases")),
    strip.position = "left"
  ) +
  labs(x = "Calendar week", y = NULL) +
  theme_who(strip.placement = "outside")
```

### Confidence intervals

The function `ci_proportion` calculates confidence intervals for proportions. In
the example below, we simulate a gender distribution of cases between different
places and plot the gender proportion with confidence intervals.

```{r, ci_proportion}

## simulate a random gender proportion for different places
genders <- tibble(
  prop_male = rbeta(10, 2, 3),
  prop_female = 1 - prop_male,
  sample = map(
    prop_male,
    ~ sample(c("male", "female"), 100, prob = c(.x, 1-.x), replace = TRUE)
  ),
  place = fct_reorder(glue("Place {LETTERS[1:10]}"), prop_male),
)

genders %>%
  mutate(
    ci = ci_proportion(
      ## calculate the number of individuals that are male
      k = map_dbl(genders$sample, ~ sum(.x == "male")),
      ## calculate the total number of individuals
      n = lengths(genders$sample)
    )
  ) %>%
  unnest_wider(ci) %>%
  ggplot(aes(place, prop_male)) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    width = 0.5
  ) +
  geom_point() +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent
  ) +
  labs(x = "Place", y = "Proportion male") +
  theme_who()
```

### Binning numbers

The `bin_numbers` function does exactly that, with the default binning set to
WHO age categories. In the example below we bin a simulated distribution of
ages and plot the distribution.

```{r, bin_numbers}
## example dataset of ages
tibble(
  age = rlnorm(100, log(30), log(2.5)),
  binned = bin_numbers(age)
) %>%
  ggplot(aes(x = binned)) +
  geom_bar() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Age category", y = "Count") +
  theme_who()
```


## Reference data

### Country dataset

The `ref_country` object contains a reference dataset for all countries. 
This has official country names, who regions, and iso2 and 3 codes for all countries globally.
Use this in joins to ensure official country names are used, and to append who region and iso codes to lists of countries. 

```{r ref_country, echo=FALSE}
gt(head(ref_country)) %>%
  tab_options(table.font.size = 12) %>%
  theme_who_gt()
```

<br> 
<br>

One example of a use cases is as follows:

```{r ref_country example}
# joining with ref_country example
tibble(
  office_country = c("Switzerland", "Germany", "Denmark", "France",
                     "Philippines", "Congo", "India", "Egypt", "United States of America")
) %>% 
  left_join(
    ref_country, by = c("office_country" = "country")
  ) %>% 
  mutate(who_region_full = relabel_who(who_region, output = "office", length = "long")) %>%
  gt() %>%
  tab_options(table.font.size = 12) %>%
  theme_who_gt()
```
### Synonyms table
The `country_synonyms` object is a table of country synonyms which can be used to convert country name synonyms to their official WHO names. The table is structured as the following, with a simple from and to column strucutre:

```{r country_synonyms, echo=FALSE}

gt(head(country_synonyms)) %>%
  tab_options(table.font.size = 12) %>%
  theme_who_gt()

```

This can be used in joins to correct country names, and ultimately join country names to metadata, as in the `ref_country` dataset. A function to automate joining of country names is also included (`convert_country_names()`). This function uses the prepackaged synonym table by default, but can be used with user-supplied reference data as well. It can be used as in the following:

```{r convert_country_names}

# dataset with a mix of official and non-official names
tibble(
  data_source = c("DRC", "Republic of Congo", "Tanzania", "France", "UK", "Germany",
                  "Vietnam", "China", "Bolivia", "Brazil")
) %>% 
  convert_country_names(country_col = "data_source", 
                        show_corrections = TRUE, 
                        ref = country_synonyms,
                        ref_from_col = "from", 
                        ref_to_col = "to"
  ) %>% 
  left_join(
    ref_country, c("data_source" = "country")
  ) %>% 
  gt() %>%
  tab_options(table.font.size = 12) %>%
  theme_who_gt()

```
