---
title: "An introduction to `whotools`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An introduction to `whotools`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4,
  warning = FALSE
)
```

## Overview

`whotools` is intended as a lightweight package for facilitating analytics at
WHO. The following vignette will cover some example use cases. We will use the
example dataset `incidence`, containing daily reported cases and deaths for a
given country:

```{r setup}
## load packages
pacman::p_load(tidyverse, lubridate, glue, whotools, gt)

## load example data from package
incidence <- whotools:::incidence
```

## Themes

`whotools` includes a theme that can be added to any ggplot2 object using the
`theme_who()` function. In the example below we are visualising the daily case
incidence curve with the WHO theme.

```{r, theme_who}
incidence %>%
  ggplot(aes(date, cases)) +
  geom_col(width = 1) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Date", y = "Daily number of cases") +
  theme_who()
```

## Palettes

The `who_palette` function provides a number of palettes that can be applied in
different scenarios and passed via the `scale_manual` option in ggplot2. In the
section below, we simulate a gender distribution of cases between different
places and plot the results using the `binary` palette.

```{r, who_palette_binary}
## simulate a random gender proportion for different places
genders <- tibble(
  prop_male = rbeta(10, 2, 3),
  prop_female = 1 - prop_male,
  sample = map(
    prop_male,
    ~ sample(c("male", "female"), 100, prob = c(.x, 1-.x), replace = TRUE)
  ),
  place = fct_reorder(glue("Place {LETTERS[1:10]}"), prop_male),
)

genders %>%
  pivot_longer(-c(place, sample)) %>%
  ggplot(aes(place, value, fill = name)) +
  geom_col(position = "stack") +
  scale_fill_manual(
    ## use binary palette type
    values = who_palette(type = "binary2"),
    labels = c(prop_female = "Female", prop_male = "Male"),
    name = NULL
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0)),
    labels = scales::percent
  ) +
  labs(x = "Place", y = "Proportion") +
  theme_who() +
  theme(legend.position = "bottom")
```

## Region labels and palettes
The `relabel_who` function helps convert regional offices labels between their
abbreviated and full version (.e.g "EMRO" and "Regional Office for the Eastern
Mediterranean") and between office and region designations (e.g. EMRO and
EMR). Below we show average case incidence per region in a table, also using the
`theme_who_gt` function, which applies a WHO theme to table outputs from the
`gt` package.

```{r, region_labels}
cases_region <- tibble(
  office = c("AFRO", "EURO", "EMRO", "PAHO", "SEARO", "WPRO"),
  region = relabel_who(office, output = "region", length = "long"),
  cases = round(runif(length(office), 20, 200), 1.1)
)

cases_region %>%
  select(-office) %>%
  gt() %>%
  cols_label(cases = "Daily cases per 100,000 population", region = "Region") %>%
  theme_who_gt()
```

We can also plot this in a bar chart and use the regional color palette from `who_palette`:

```{r, region_palette}
cases_region %>%
  mutate(office = fct_reorder(office, cases)) %>%
  ggplot(aes(office, cases, fill = office)) +
  geom_col() +
  scale_x_discrete(labels = ~ relabel_who(.x, output = "region", length = "short")) +
  scale_fill_manual(values = who_palette(type = "region"), guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "WHO Region", y = "Daily cases per 100,000 population") +
  theme_who()
```

## Moving averages

The `moving_average` function calculates a moving average across a specified
date window. Below we add a line showing the average daily incidence over a
sliding 7-day window.

```{r, moving_average}
incidence %>%
  mutate(
    cases_weekly = moving_average(val = cases, date_val = date, before = 3, after = 3),
  ) %>%
  ggplot(aes(x = date)) +
  geom_col(aes(y = cases), width = 1) +
  geom_line(aes(y = cases_weekly), size = 2, color = "black") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Date", y = "Daily number of cases") +
  theme_who()
```

## Percent changes

The `perc_change` function will calculate percent changes. We can use this to
visualise percent changes in weekly reported cases:

```{r, perc_change}
incidence %>%
  group_by(week = isoweek(date)) %>%
  filter(week < 30) %>%
  summarise(cases = sum(cases)) %>%
  mutate(change = 100*perc_change(cases, lag(cases))) %>%
  pivot_longer(-week) %>%
  ggplot(aes(week, value)) +
  geom_col() +
  facet_wrap(
    ~ name,
    ncol = 1,
    scales = "free_y",
    labeller = labeller(name = c(cases = "Weekly number\nof cases", change = "% change in\nweekly cases")),
    strip.position = "left"
  ) +
  labs(x = "Calendar week", y = NULL) +
  theme_who(strip.placement = "outside")
```

## Confidence intervals

The function `ci_proportion` calculates confidence intervals for proportions. We
can plot the gender proportion plot above in a different way to include
confidence intervals using this function. 

```{r, ci_proportion}
genders %>%
  mutate(
    ci = ci_proportion(
      ## calculate the number of individuals that are male
      k = map_dbl(genders$sample, ~ sum(.x == "male")),
      ## calculate the total number of individuals
      n = lengths(genders$sample)
    )
  ) %>%
  unnest_wider(ci) %>%
  ggplot(aes(place, prop_male)) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    width = 0.5
  ) +
  geom_point() +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent
  ) +
  labs(x = "Place", y = "Proportion male") +
  theme_who()
```

## Binning numbers

The `bin_numbers` function does exactly that, with the default binning set to
WHO age categories. In the example below we bin a simulated distribution of
ages and plot the distribution.

```{r, bin_numbers}
## example dataset of ages
tibble(
  age = rlnorm(100, log(30), log(2.5)),
  binned = bin_numbers(age)
) %>%
  ggplot(aes(x = binned)) +
  geom_bar() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Age category", y = "Count") +
  theme_who()
```


## Reference data

The `ref_country` object contains a reference dataset for all countries. 
This has official country names, who regions, and iso2 and 3 codes for all countries globally.
Use this in joins to ensure official country names are used, and to append who region and iso codes to lists of countries. 

```{r ref_country, echo=FALSE}

knitr::kable(head(ref_country))

```


<br> 
<br>

One example of a use cases is as follows:

```{r ref_country example}
# joining with ref_country example
tibble(
  office_country = c("Switzerland", "Germany", "Denmark", "France",
                     "Philippines", "Congo", "India", "Egypt", "United States of America")
) %>% 
  left_join(
    ref_country, by = c("office_country" = "country")
  ) %>% 
  mutate(who_region_full = relabel_who(who_region, output = "office", length = "long")) %>% 
  knitr::kable()
```
